---
title: "ESM 244: Lizards Model"
description: |
  Classwork for ESM 244 (UCSB) non least squares regression.
author:
  - name: Kristina Fauss
date: 2022-02-21
output:
  html_document:
    code_folding: hide
  distill::distill_article:
    self_contained: false
---

```{r setup, include=TRUE, message= FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning = FALSE)
library(tidyverse)
library(Metrics)
library(cowplot)
library(knitr)
library(sjPlot)

```

## Overview

### Data:
Title: 
Lizard pitfall trap data from 11 NPP study locations at the Jornada Basin LTER site, 1989-2006.

Description: 
The time period captured is 1989 to 2006. The data includes the following contents as information about lizards captured at the Jornada Basin LTER across 11 study locations: [date, zone, site, plot, pit, spp, sex, rcap, toe_num, SV_length, weight, tail, pc]. 

Site, plot, and pit refer to capture location, while zone refers to vegetation. Spp indicates a species code. Rcap indicates whether the lizard is a recapture. Sex, toe_num, SV_length, total_length, weight, and tail all reference characteristics of the lizard. Data is intended to track how desertification (change in vegetation) in the Chihuahuan desert affects consumer species. 

Citation: 
*Lightfoot, D. and W.G. Whitford. 2020. Lizard pitfall trap data from 11 NPP study locations at the Jornada Basin LTER site, 1989-2006 ver 37. Environmental Data Initiative. https://doi.org/10.6073/pasta/4a6e258fb49c31e222ecbbcfd128967f (Accessed 2022-02-03).*

```{r}
# Read in cleaned data:
liz_data <- read.csv('lizard.csv')
```

### Questions:
Apply non least squares regression to estimate model parameters for a lizard length to weight model.

### Data Exploration
```{r}
ggplot(data=liz_data, aes(x=SV_length, y=weight))+
  labs(title='Length v. Weight Across Species',x='Length',y='Weight',color='Species')+
  geom_point(aes(color=spp,shape=sex))
```

### Fit a Snout Length to Weight Model

Original Model:
\[
W = a(SVL)^b
\]

Log transform:
\[
ln(W) = b*ln(SVL) + ln(a)
\]

```{r}
# Calculate log of SVL and Weight
liz_data$log_weight <- log(liz_data$weight)
liz_data$log_SV_length <- log(liz_data$SV_length)
```

OLS Model of Log-Transformed Data: Supply coeficients to NLS Model
```{r}
OLSmod_liz <- lm(log_weight ~ log_SV_length, data=liz_data)

# Get coefficients
coefs_liz <- as.data.frame(coef(OLSmod_liz))
b = coefs_liz$`coef(OLSmod_liz)`[2]
log_a = coefs_liz$`coef(OLSmod_liz)`[1]
a = exp(log_a)
```

Define original function:
```{r}
f_liz<-function(coef_a,coef_b,SV_length){
  out=coef_a*(SV_length^coef_b)
  return(out)
}
```

NLS model:
```{r}


# Guess carrying capacity as the largest val in our data
guess_vec=c(a,b)

NLSmod_liz=nls(weight~f_liz(coef_a,coef_b,SV_length),
                  data=liz_data,
                  start=list(coef_a=guess_vec[1],coef_b=guess_vec[2]),
                  trace=TRUE) # reached a convergence quickly?

tab_model(NLSmod_liz)

```

Plot the NLS model and data:
```{r}
# Predict data from NLS model:
liz_data$weight_pred<-predict(NLSmod_liz)

# Plot
ggplot()+
  geom_point(data=liz_data, aes(x=SV_length, y=weight_pred, color=sex))+
  labs(title='Length v. Weight Across Species',x='Length',y='Weight',color='Sex')
```

### Western Whiptail Lizard Model:
```{r}
# Filter out Western Whiptail Lizard Data: CNTI
cnti_data <- filter(liz_data,spp=='CNTI',sex=='M')
```

Data Exploration, CNTI

```{r}
ggplot(data=cnti_data, aes(x=SV_length, y=weight))+
  labs(title='Length v. Weight Western Whiptail',x='Length',y='Weight',shape='Sex')+
  geom_point(aes(shape=sex))
```

### Fit the Snout Length to Weight Model on CNTI

```{r}
# Calculate log of SVL and Weight
cnti_data$log_weight <- log(cnti_data$weight)
cnti_data$log_SV_length <- log(cnti_data$SV_length)
```

OLS Model of Log-Transformed Data: Supply coeficients to NLS Model
```{r}
OLSmod_cnti <- lm(log_weight ~ log_SV_length, data=cnti_data)

# Get coefficients
coefs_cnti <- as.data.frame(coef(OLSmod_cnti))
b_cnti = coefs_cnti$`coef(OLSmod_cnti)`[2]
log_a_cnti = coefs_cnti$`coef(OLSmod_cnti)`[1]
a_cnti = exp(log_a_cnti)
```

Define original function:
```{r}
f_cnti<-function(coef_a,coef_b,SV_length){
  out=coef_a*(SV_length^coef_b)
  return(out)
}
```

NLS model:
```{r}


# Guess carrying capacity as the largest val in our data
guess_vec_cnti=c(a_cnti,b_cnti)

NLSmod_cnti=nls(weight~f_cnti(coef_a,coef_b,SV_length),
                  data=cnti_data,
                  start=list(coef_a=guess_vec_cnti[1],coef_b=guess_vec_cnti[2]),
                  trace=TRUE) # reached a convergence quickly?

tab_model(NLSmod_cnti)
```

Plot the NLS model and data:
```{r}
# Predict data from NLS model:
cnti_data$weight_pred<-predict(NLSmod_cnti)

# Plot
ggplot()+
  geom_point(data=cnti_data, aes(x=SV_length, y=weight_pred, color=sex))+
  labs(title='Length v. Weight Across Species',x='Length',y='Weight',color='Sex')
```
### Plot of both models:

```{r}
# Filter model output trained on total to just CNTI for comparison:
liz_data_cnti<-filter(liz_data,spp=='CNTI',sex=='M')

# Calculate RMSE's
tot_mod_rmse <- rmse(liz_data_cnti$weight,liz_data_cnti$weight_pred)
cnti_mod_rmse <- rmse(cnti_data$weight,cnti_data$weight_pred)

cap=paste('RMSE for Total model over CNTI:',round(tot_mod_rmse,2),'\nRMSE for CNTI model over CNTI:',round(cnti_mod_rmse,2),'\nThe Male-CNTI specific model is better at predicting weights of Male CNTI lizards \nand should be used for this specific case but may not generalize well to other species.')
```


```{r}
# Plot
ggplot()+
  geom_point(data=cnti_data, aes(x=SV_length, y=weight_pred, col='CNTI'))+
  geom_point(data=liz_data_cnti, aes(x=SV_length, y=weight_pred, col='Total'))+
  labs(title='Length v. Weight Across Species',x='Length',y='Weight',color='Model',caption=cap)
```




